Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
# edited isotopic data methods

\
\    To account for non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

@@

To validate the ERD models we used data for xylem sap $\mathrm{\delta}$^2^H (‰) as an independent observation of ERD. We used xylem sap $\mathrm{\delta}$^2^H values given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil (Fig. S\@ref(fig:iso-data)). Briefly, this method makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground. Generally due to evaporative enrichment (here, during the dry season), shallow soil water has a greater deuterium to hydrogen content than deep water (here, $\mathrm{\delta}$^2^H ranging from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water. As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H of water in the xylem sap of a tree reflects signature of the source water, and the tree's water-sourcing depth can therefore be identified as the soil depth with the same value of $\mathrm{\delta}$^2^H as of the Xylem sap, or a mixture of shallower or deeper depths [@Dawson.1991]. Analyses of xylem sap $\mathrm{\delta}$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water. Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and soil water $\mathrm{\delta}$^2^H was variable in the 0.2 to 0.7 m depth range.

\
\   We compared six species with data for $\mathrm{\delta}$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves. We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile. For each model of ERD (Eq. \@ref(eq:M-chosen) and Eqs. in SI), estimated species ERD was regressed against species xylem sap $\mathrm{\delta}$^2^H (‰).

>>>>

\
\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

@@

Xylem sap $\mathrm{\delta}$^2^H values can provide an index of rooting depth, which we used to validate the ERD models as an independent observation of ERD. We used xylem sap $\mathrm{\delta}$^2^H values from BCI given in @Meinzer:2001 for the dry-season of March 1997 as this period showed largest seasonal divergence among species, with a corresponding vertical gradient of $\mathrm{\delta}$^2^H in the soil. (Fig. S\@ref(fig:iso-data)). Briefly, this isotopic approach makes use of a naturally formed gradient in deuterium to hydrogen ratio of soil water vertically belowground, which is generally formed due to evaporative enrichment (here, during the dry season). Shallow soil water has a greater deuterium to hydrogen content than deep soil water (here, $\mathrm{\delta}$^2^H ranged from -16.5 to -50 ‰ over 0.05 to 1 m depth), with ground-water having the lowest deuterium to hydrogen content (here, -60 ‰), typically reflecting un-fractionated rain water. As root water-uptake is generally a non-fractionating process, $\mathrm{\delta}$^2^H of water in the xylem sap of a tree reflects the signature of the source water. Thus water-sourcing depth of the tree can be identified as the depth with the same value of $\mathrm{\delta}$^2^H as that of the xylem sap, with the uncertainty that the value of $\mathrm{\delta}$^2^H may also result from a mixing of water from shallower or deeper depths [@Dawson.1991]. Analyses of xylem sap $\mathrm{\delta}$^2^H of the 12 species studied by @Meinzer:2001 showed that four species sourced water between 1 m and ground water. Two other had water access at least 0.8 m deep, one was restricted to water access between 0 to 0.2 m and five fell in the range over which the relationship between soil depth and soil water $\mathrm{\delta}$^2^H was variable in the 0.2 to 0.7 m depth range.

\   We compared six species with data for $\mathrm{\delta}$^2^H as well as modelled estimates of ERD, after removing those species in @Meinzer:2001 that are likely to have no leaves in March 1997 because xylem sap $\mathrm{\delta}$^2^H of these species may not be linked with water sourced at the time of measurement and thus may not be comparable with other species that had leaves. We also removed *Guapira standleyana*, whose xylem sap isotope value was not well constrained by the soil water isotope depth profile. For each model of ERD (Eq. \@ref(eq:M-chosen) and Eqs. in SI), estimated species ERD was regressed against species xylem sap $\mathrm{\delta}$^2^H (‰).
